The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2016-07-23 article.tex
Some last cleanup for today: figure numbering (latex approach doesn't work), some reshuffling)

\author[
   addressref={um},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={egon.willighagen@maastrichtuniversity.nl}   % email address
]{\inits{ELW}\fnm{Egon L} \snm{Willighagen}}
\author[
   addressref={aff2},
   email={john@nextmovesoftware.com}
]{\inits{JWM}\fnm{John W} \snm{May}}
\author[
   addressref={uppsala},
   email={jonathan.alvarsson@farmbio.uu.se}
]{\inits{JA}\fnm{Jonathan} \snm{Alvarsson}}
\author[
   addressref={uppsala},
   email={berg.arvid@gmail.com}
]{\inits{AB}\fnm{Arvid} \snm{Berg}}
\author[
   addressref={idea},
   email={jeliazkova.nina@gmail.com}
]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
\author[
   addressref={???},
   email={miguelrojasch@googlemail.com}
]{\inits{MRJ}\fnm{Miguel}~\snm{Rojas-Cherto}}
\author[
   addressref={???},
   email={pluskal@wi.mit.edu}
]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
\author[
   addressref={uppsala},
   email={ola.spjuth@farmbio.uu.se}
]{\inits{OS}\fnm{Ola} \snm{Spjuth}}
\author[
   addressref={???},
   email={gilleain.torrance@gmail.com}
]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
\author[
   addressref={nih},
   email={guhar@mail.nih.gov}
]{\inits{RG}\fnm{Rajarshi} \snm{Guha}}
\author[
   addressref={ebi},
   email={steinbeck@ebi.ac.uk}
]{\inits{CS}\fnm{Christoph} \snm{Steinbeck}}

@@

  This CDK version has isolated a new atom typing framework, removing the perception of
  atom types from various algorithms, allowing the perception to be tested separately.
  The new code defines the atom types using a list that specifies for each type
  the element symbol, hybridization, formal charge, number of lone pairs, and
  an enumeration of the bond orders. This list of properties captures the
  information needed for the various algorithms in the CDK. For example,
  hybridization information can be used in certain aromaticity models (see later),
  and the lone pair information is needed for resonance structure calculation
  needed, for example, for Gasteiger $\pi$-charges.

@@

  Many of the rendering options are available as parameters in the core API
  and as methods on the DepictionGenerator class. This includes substructure
  coloring, exemplified with this example reaction (see Fig.~\ref{fig:depiction}):

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. Particularly, collision detection has been
  greatly improved. Figure~\ref{fig:sdg} shows a difference in output between
  the old code base, with and without overlap resolving, and with the new code.
  The new code can be used like:

@@

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of that class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~\ref{tab:fingerprints}.

@@

... describe all modules, maybe add graphviz plot ...

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~\ref{tab:modules} and the dependencies
between the CDK modules are depicted in Figure~\ref{fig:deps}.

@@

%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \subsection*{Figure 1 - Dependencies between CDK modules.}\label{fig:deps}
      Visualization of the dependencies between CDK modules. For example,
      the cdk-core depends on the cdk-interfaces module. A few higher level
      modules have been left out: cdk-builder3dtools, cdk-legacy, and
      cdk-depict.

\subsection*{Figure 2 - Atom type information specified for a sp3-hybridized
carbon.}\label{fig:atomtype}
\begin{verbatim}
  <at:AtomType rdf:ID="C.sp3">
    <at:categorizedAs rdf:resource="&cdkat;C.tetrahedral"/>
    <at:hasElement rdf:resource="&elem;C"/>
    <at:hybridization rdf:resource="&at;sp3"/>
    <at:formalCharge>0</at:formalCharge>
    <at:lonePairCount>0</at:lonePairCount>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
  </at:AtomType>
\end{verbatim}

@@

\subsection*{Figure 4 - Integrated example showing the rendering and SMILES
parsing functionality.}\label{fig:depiction}

@@

  \subsection*{Table 1 - A selection of key CDK modules with major changes.}\label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Dependencies between modules are depicted in Figure~\ref{fig:deps}.
  \vskip 1\baselineskip

@@

>>>>

\author[
   addressref={um},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={egon.willighagen@maastrichtuniversity.nl}   % email address
]{\inits{ELW}\fnm{Egon L} \snm{Willighagen}}
\author[
   addressref={aff2},
   email={john@nextmovesoftware.com}
]{\inits{JWM}\fnm{John W} \snm{May}}
\author[
   addressref={uppsala},
   email={jonathan.alvarsson@farmbio.uu.se}
]{\inits{JA}\fnm{Jonathan} \snm{Alvarsson}}
\author[
   addressref={uppsala},
   email={berg.arvid@gmail.com}
]{\inits{AB}\fnm{Arvid} \snm{Berg}}
\author[
   addressref={idea},
   email={jeliazkova.nina@gmail.com}
]{\inits{NJ}\fnm{Nina} \snm{Jeliazkova}}
\author[
   addressref={???},
   email={pluskal@wi.mit.edu}
]{\inits{TP}\fnm{Tomáš}~\snm{Pluskal}}
\author[
   addressref={???},
   email={miguelrojasch@googlemail.com}
]{\inits{MRJ}\fnm{Miguel}~\snm{Rojas-Cherto}}
\author[
   addressref={uppsala},
   email={ola.spjuth@farmbio.uu.se}
]{\inits{OS}\fnm{Ola} \snm{Spjuth}}
\author[
   addressref={???},
   email={gilleain.torrance@gmail.com}
]{\inits{GT}\fnm{Gilleain} \snm{Torrance}}
\author[
   addressref={nih},
   email={guhar@mail.nih.gov}
]{\inits{RG}\fnm{Rajarshi} \snm{Guha}}
\author[
   addressref={ebi},
   email={steinbeck@ebi.ac.uk}
]{\inits{CS}\fnm{Christoph}~\snm{Steinbeck}}

@@

  This CDK version has isolated a new atom typing framework, removing the perception of
  atom types from various algorithms, allowing the perception to be tested separately.
  The new code defines the atom types using a list that specifies for each type
  the element symbol, hybridization, formal charge, number of lone pairs, and
  an enumeration of the bond orders (see Figure~1). This list of properties captures the
  information needed for the various algorithms in the CDK. For example,
  hybridization information can be used in certain aromaticity models (see later),
  and the lone pair information is needed for resonance structure calculation
  needed, for example, for Gasteiger $\pi$-charges.

@@

  Many of the rendering options are available as parameters in the core API
  and as methods on the DepictionGenerator class. This includes substructure
  coloring, exemplified with this example reaction (see Fig.~2\ref{fig:depiction}):

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. Particularly, collision detection has been
  greatly improved. Figure~3\ref{fig:sdg} shows a difference in output between
  the old code base, with and without overlap resolving, and with the new code.
  The new code can be used like:

@@

\subsubsection*{Molecular fingerprints}
Also the molecular fingerprints in CDK has seen development since the 1.2
release. Traditionally CDK fingerprints were represented using the
\textit{BitSet} class shipping with Java. The benefits of that class include
such things as already implemented methods for modifying bit sets using other
bit sets. However the class keeps a vector of bits in memory. The solution was
excellent for hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a fingerprint
designed to avoid collisions with, \textit{e.g.} a 32 bit indexing space using
this approach would be highly memory-inefficient. To allow for multiple
fingerprint representation implementations a bit fingerprint interface was
introduced. Also, although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer vectors
supported in CDK as well. The fingerprints currently existing in CDK are listed
in Table~1\ref{tab:fingerprints}.

@@

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~2\ref{tab:modules} and the dependencies
between the CDK modules are depicted in Figure~4\ref{fig:deps}.

@@

\section*{Figures}

\subsection*{Figure 1 - Atom type information specified for a sp3-hybridized
carbon.}\label{fig:atomtype}
\begin{verbatim}
  <at:AtomType rdf:ID="C.sp3">
    <at:categorizedAs rdf:resource="&cdkat;C.tetrahedral"/>
    <at:hasElement rdf:resource="&elem;C"/>
    <at:hybridization rdf:resource="&at;sp3"/>
    <at:formalCharge>0</at:formalCharge>
    <at:lonePairCount>0</at:lonePairCount>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
    <at:formalBondType rdf:resource="&bo;single"/>
  </at:AtomType>
\end{verbatim}

\subsection*{Figure 2 - Integrated example showing the rendering and SMILES
parsing functionality.}\label{fig:depiction}

@@

  \subsection*{Figure 4 - Dependencies between CDK modules.}\label{fig:deps}
      Visualization of the dependencies between CDK modules. For example,
      the cdk-core depends on the cdk-interfaces module. A few higher level
      modules have been left out: cdk-builder3dtools, cdk-legacy, and
      cdk-depict.

@@

@@

      \subsection*{Table 2 - A selection of key CDK modules with major changes.}\label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Dependencies between modules are depicted in Figure~4\ref{fig:deps}.
  \vskip 1\baselineskip
